Benchmarking of signaling networks generated by large language models

Version 1

Abstract

Computational models of signaling networks provide frameworks for predicting how molecular cues guide cell decisions. But they are typically limited by manual curation from incomplete literature. Here, we test whether general-purpose large language models (LLMs) generate accurate models of signaling networks. We find that general purpose LLMs generate 24-58% of the reactions of literature-curated networks for cardiomyocyte hypertrophy, myofibroblast activation, and mechano-signaling, and predicting network responses to perturbations with accuracies of 5-26%. While current general-purpose LLMs generate signaling networks with limited accuracy, this study provides a pipeline and benchmarks to guide future improvements.

Main

The response of cells to external stimuli is a fundamental characteristic of life, mediated by complex molecular networks of signal transduction and gene regulation. Computational models of signaling networks have been widely used to understand their systems properties and new guide experiments1,2. However, a central hurdle is that computational models of signaling networks typically require substantial manual literature curation of network reactions and rate constants, which is time intensive and requires extensive experimental knowledge3. Further, existing data is often incomplete.

AI, and large language models (LLMs) in particular, have the potential to outperform human labor-intensive processes of knowledge curation. For example, general purpose LLMs have scored highly on the U.S. Medical Licensing Exam4, law exams5, and Advanced Placement exams for many subjects5. AI has also made substantial contributions in biology, including predicting protein structure6,7, cell types8,9, and is expected to play a central role in the development of comprehensive virtual cells10–12. However, to our knowledge, LLM-generated predictive models of signaling networks have not been reported or benchmarked. Here, we develop a pipeline for LLM-generated models of signaling networks, and we test whether general purpose LLMs generate networks with accurate structure and predictions of perturbation responses, as benchmarked against three manually-curated and validated models13–15.

We first established a computational pipeline to generate network models from LLMs using a given gene set (Figure 1A). For benchmarking purposes, here the gene set is based on annotations of manually curated network models that have been extensively validated. However, future studies could derive gene sets from other sources such as proteomics or transcriptomics. We optimized a programmatic LLM prompt that iteratively queried for directed (activation or inhibition) interactions among the gene products in the context of a particular phenotype (here “hypertrophy”, “fibroblast”, or “mechanosignaling”). With text processing, we extracted the returned directed interactions to examine network structure and then converted the network into the format needed to simulate network responses to perturbations, here with logic- based differential equations in Netflux16. LLM-based models were generated 10 times per network to examine reproducibility.

Signaling networks generated by general-purpose large language models.

A) Schematic of pipeline for LLM-generated models of signaling networks. B) Network reactions recalled by three large language models (Gemini2.0, orange; ChatGPT4, blue; Claude3.7, green) compared with a “Ground Truth” literature-curated and validated cardiomyocyte hypertrophy signaling network13 (gray reactions). LLM networks were generated using iterative prompts based on the gene set list of the Ground Truth hypertrophy network. Summary of reaction recall accuracy for three literature-curated signaling networks (C, hypertrophy13; D, fibroblast14; and E, mechanosignaling15) by Gemini, GPT, and Claude. * Indicates p < 10−9 in one-sample T test between LLM-generated replicates (n = 10) and the ground truth network.

We first tested the extent to which three general purpose LLMs (GPT 4.1, Open AI; Gemini 2.0, Google; and Claude 3.7, Anthropic) could reconstruct the structure of the manually-curated cardiomyocyte hypertrophy signaling network13, based on its gene annotations alone. Overall, all three LLMs capture aspects of this manually-curated network, but with varying coverage depending on the region of the network and the particular LLM. In particular, GPT 4.1, Gemini 2.0, and Claude 3.7 all performed fairly well at reconstructing upstream aspects of ligand-receptor interactions (e.g. EGF to EGFR) and well- established second messenger cascades like β-adrenergic receptor to cAMP to CREB pathway or phospholipase C to diacylglycerol to PKC. They also accurately predicted the structure of highly conserved signaling axes such as PI3K-Akt-mTor and calcium- calmodulin-calcineurin-NFAT. In contrast, the three LLMs were less able to reconstruct regulation of downstream transcription factors (e.g. cFOS, NFAT, cJUN, MEF2) and gene expression (SERCA, αMHC, βMHC, natriuretic peptides ANP/BNP) or cell area / hypertrophy. The lower reconstruction accuracy of downstream signaling may arise because these contain more cardiomyocyte-specific reactions, but such gaps could also represent a general challenge of LLMs in reconstructing gene regulation from literature. Overall, the three LLMs exhibited 26.70-58.12% reconstruction of network reactions in the cardiomyocyte hypertrophy network. The top-performing LLM was Claude 3.7, which was reproducible with repeated queries.

To what extent is the limited recall of LLM-generated networks specific to cardiomyocyte hypertrophy? To address this question, we used GPT 4.1, Gemini 2.0, and Claude 4.7 to also reconstruct network models of fibroblast activation14 (91 nodes, 134 reactions) (Supplementary Figure 1) and mechanosignaling15 (94 nodes, 125 reactions) (Supplementary Figure S2). Like the hypertrophy model, these networks were previously manually reconstructed from the literature, translated into predictive logic-based models, and had their predictions validated against substantial experiments from the literature that were not used to develop the model. Consistent with the hypertrophy network, LLM-generated models of fibroblast activation and mechanosignaling successfully reconstructed 25.93-50.00% and 24.42-48.26% of their manually curated counterparts. Across all three networks, Claude 4.7 and GPT 4.1 outperformed Gemini 2.0, but with limited recall of the overall network.

We then asked whether the regional coverage of LLM-generated networks of fibroblasts and mechanosignaling mirrors that of cardiomyocytes. Indeed, the LLM- generated fibroblast network was most effective at reconstructing ligand-receptor interactions and highly conserved pathways such as cAMP and Ras/Raf/ERK, but could not recall AT1R signaling, PDGFR-abl signaling, and especially downstream regulation of transcription factors, genes, and extracellular matrix (e.g. MMP-collagen subnetwork) (Supplementary Figure 1). Unlike hypertrophy and fibroblast networks, mechanosignaling is mediated by stretch-dependent proteins that are relatively less characterized. Indeed, the LLM-generated networks for mechanosignaling did not adequately reconstruct direct mechano-sensation and had limited coverage of gene regulation consistent with the other networks (Supplementary Figure 2). Still, core conserved pathways for Na/Ca regulation, regulation of protein synthesis, and MAPK signaling were reconstructed well.

Models of signaling networks are useful not just for their description of structure, but especially for their ability to simulate the response to new perturbations. Making such predictions requires a specific mathematical formalism. For the three networks used for benchmarking, we had previously automatically translated them into logic- based differential equation models17 using the software Netflux16. This allows network- wide prediction of dynamics in response to any desired perturbations and does not require prior knowledge of rate constants or concentrations17. The predictions of these three logic-based models have also been highly validated against experiments not used to develop the model13–15,18. Therefore, we generated logic-based models based on all of the LLM-generated network structures described above.

Returning our focus to the hypertrophy network, we examined the ability of LLM- generated networks to predict classic the “fetal gene program” gene expression signature in cardiomyocytes. The manually curated model accurately predicts that AngII and/or ISO increase the expression of ANP, BNP, GATA4, βMHC, sACT, and increases cell area/size, while decreasing expression of SERCA and αMHC. In contrast, the LLM- generated network models produce limited aspects of this signature, with the best performance in predicting ISO-dependent ANP in the GPT version (Figure 2A). We then tested whether this validation accuracy scales across all available manually curated experiments that cover a wider array of perturbations and measured outputs. Overall, for 114 validations where the manually curated network has an functional accuracy of 94.74%, LLM-generated networks have an accuracy of 6.14-26.32% (Figure 2B).

Experimental validation of perturbation responses predicted by LLM-generated signaling network models.

A) Representative validations of network models generated by manual curation or by LLMs (Gemini, GPT, Claude), in comparison to experiments in conditions of Angiotensin II (AngII) or isoproterenol (ISO) from the literature13. B) Summary of systematic validations of manually curated (Ground Truth) and LLM-generated network models of hypertrophy, fibroblast, and mechano-signaling network models against perturbation experiments from the literature (n = 114, 83, and 171 experiments, respectively). * Indicates p < 10−11 in one-sample T test between LLM-generated model validation scores (n = 10 replicates) and ground truth model validation accuracy.

Repeating this systematic experimental validation for LLM-generated models of the fibroblast (Figure 2C) and mechano-signaling (Figure 2D), we observed similar validation accuracies of 16.87-21.69% and 5.81%, respectively. Together, these results indicate that validation accuracy of LLM-generated models for perturbation responses is even lower than their accuracy for reconstructing network structure. Because the validation tests are testing larger-range functional interactions, this difference may be caused by gaps in reconstructed pathways.

In conclusion, we developed a pipeline for generating predictive, logic-based models of signaling networks using general purpose large language models (LLMs). To test the reconstruction and simulation accuracy of LLM-generated network models, we benchmarked this pipeline using three LLMs (GPT 4.1, Claude 4.7, and Gemini 2.0) against three large-scale, manually curated and validated models of signaling networks. Across cardiac hypertrophy, fibroblast, and mechano-signaling networks, LLM- generated models exhibit 24.42-58.12% reconstruction of reactions from the manually curated models. Reconstruction accuracy is high for ligand-receptor interactions and well-conserved pathways, but lower for cell-specific gene regulation or biophysical regulation. Likely due to gaps in reconstructed pathways, LLM-generated models had a low ability to accurately simulate the responses of networks to perturbations. Overall, we provide a pipeline for LLM-generated models of signaling networks and benchmark their performance against three manually curated networks. Improving the performance of LLM-generated network models may require hybrid manual-LLM approaches or advances in more special-purpose AI models.

Methods

Prior knowledge models of signaling networks

To benchmark LLM-generated network models, we selected three large-scale curated network models based on manual curation of the directed interactions using the literature13–15. These models were recently used to benchmark protein interaction databases19, creating a comparison’ for the current analysis. These models contain a range of biological processes including ligand-receptor interactions, phosphorylation, gene regulation, and cellular phenotypes. Associated with the network structure is an annotated list of genes associated with the nodes in the model. These models were mathematically formulated as systems of logic-based differential equations17, which enable dynamic network-wide predictions of how those cells respond to new perturbations. The predictive accuracy of these models has previously been rigorously tested with manual curation of literature not used to develop the models, with accuracies of 94.74% (hypertrophy), 81.93% (fibroblast), and 77.33% (mechanosignaling)13–15.

Querying large language models

We focused on widely used, general-purpose large language models (LLMs) that are most relevant to wider use in the field. Specifically, we selected current versions of GPT (4.1, OpenAI), Gemini (2.0, Google), and Claude (3.7, Anthromorphic). The prompt consisted of three steps:

The list of gene symbols, along with the description: “List of genes and other signaling nodes:”.

Description of the desired outputs: “For the first {batch_size} entries in this list of genes, proteins, and other signaling nodes from a {phenotype} network, please provide more than 0 but fewer than {max_connections} direct interactions with other nodes from the list supported by available literature. Simply list the input node, affected node, and if the affected node is stimulated / inhibited.” Batch size was set at 20 genes, max connections set to the maximum number of connections exhibited in that parent network, and phenotype set at “cardiac hypertrophy”, “fibroblast”, or “mechanosignaling” corresponding to the three networks used for benchmarking.

Instructions to repeat the process until reaching the end of the gene list: “That looks great! Please do the same operation for the next {len(chunk)} nodes! Thank you”

The responses were stored as .txt files and then programmatically extracted using regular expressions.

Extracting reactions from LLM outputs, logic-based model generation

Each of the LLMs returned network reactions in slightly different format. But for example, Claude 3.7 returned reactions in the form: “ADRB2 stimulates RAC1” or “RAC1 stimulates MAP3K1”. Relationships based on directional signed relationships were kept, such as terms “stimulates”, down-regulates”, or “generates”. Returned reactions referring to non-relevant proteins or processes were discarded, as were non- directional relationships. We attempted also querying PubMed IDs, but these were rarely accurate.

Reactions successfully extracted from the LLM outputs were formatted for uniform comparison to the reaction list from the corresponding manually curated network. Finally, this reaction list was translated into the Netflux format for automated conversion to a logic-based differential equation model16,17. This final logic-based model was used to perform validation simulations.

eLife Assessment

The authors address a hard question and propose a pipeline for using Large Language Models to reconstruct signalling networks as well as to benchmark future models. The findings are valuable for a defined subfield, as the proposed framework allows for assessing such approaches systematically. The overall support is solid, although the present evaluation remains limited in scope and would benefit from a wider range of networks and performance metrics.

Reviewer #1 (Public review):

Summary:

Large language models (LLMs) have been developed rapidly in recent years and are already contributing to progress across scientific fields. The manuscript tries to address a specific question: whether LLMs can accurately infer signaling networks from gene lists. However, the evaluation is inadequate due to four major weaknesses described below. Despite these limitations, the authors conclude that current general-purpose LLMs lack adequate accuracy, which is already widely recognized. Its key contribution should instead be to provide concrete recommendations for the development of specialized LLMs for this task, which is completely absent. Developing such specific LLMs would be highly valuable, as they could substantially reduce the time required by researchers to analyze signaling networks.

Strengths:

The manuscript raises a good question: whether current LLMs can accurately generate signaling networks from gene lists.

Weaknesses:

(1) The authors evaluate LLM performance using only three signaling networks: "hypertrophy", "fibroblast", and "mechanosignaling". Given the large number of well-established signaling pathways available, this is not a comprehensive assessment. Moreover, the analysis need not be restricted to signaling networks. Other network types, including metabolic and transcriptional regulatory networks, are already accessible in well-known databases such as KEGG, Reactome, BioCyc, WikiPathways, and Pathway Commons. Including these additional networks would substantially strengthen the evaluation.

(2) In LLM evaluation, the authors use the gene lists that exactly match those in their "ground truth" networks, thereby fixing the set of nodes and evaluating only the predicted edges. However, in practical research, the relevant genes or nodes are not fully known. A more realistic assessment would therefore include gene lists with both genes present in the ground-truth network and additional genes absent from it, to evaluate the ability of the LLM to exclude irrelevant genes.

(3) The authors report only the recall/sensitivity of the LLM, without assessing specificity. In practical applications, if an LLM generates a large number of incorrect interactions that greatly exceed the correct ones, researchers may be misled or may lose confidence in the LLM output. Therefore, a comprehensive evaluation must include both sensitivity and specificity. Furthermore, it would be informative to check whether some of the "false positives" might in fact represent biologically plausible interactions that are absent from the manually curated "ground truth". Manually generated "ground truth" can overlook genuine interactions, and the ability of LLMs to recover such missing edges could be particularly valuable. This may even represent one of the most important potential contributions of LLMs.

(4) It is widely known that applying differential equation models to highly complex biological networks, such as the three networks in the manuscript, is meaningless, because these systems involve a large number of parameters whose values can drastically alter the results. As Richard Feynman once said: "with four parameters I can fit an elephant, and with five I can make him wiggle his trunk." Thus, the evaluation of LLMs on "logic-based differential equation models" does not make much sense.

Reviewer #2 (Public review):

Summary:

The authors evaluate whether commonly used LLMs (ChatGPT, Claude and Gemini) can reconstruct signalling networks and predict effects of network perturbations, and propose a pipeline for benchmarking future models. Across three phenotypes (hypertrophy, fibroblast signalling, and mechanosignalling), LLMs capture upstream ligand-receptor interactions and conserved crosstalk but fail to recover downstream transcriptional programmes. Logic-based simulations show that LLM-derived networks underperform compared to manually curated models. The authors also propose that their pipeline can be used for benchmarking future models aimed at reconstructing signalling networks.

Strength:

The authors compare the outcomes from three LLMs with three manually curated and validated models. Additionally, they have investigated gene network reconstruction in the context of three distinct phenotypes. Using logic-based modelling, the authors assessed how LLM-derived networks predict perturbation effects, providing functional validation beyond network overlap.

Weaknesses:

The authors have used legacy models for all three LLMs, and the study would benefit from testing the current versions of the LLMs (ChatGPT 5.2, Claude 4.5 and Gemini 2.5). Additional metrics such as node coverage, node invention, direction accuracy and sign accuracy would be useful to make robust comparisons across models.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Summary

Large language models (LLMs) have been developed rapidly in recent years and are already contributing to progress across scientific fields. The manuscript tries to address a specific question: whether LLMs can accurately infer signaling networks from gene lists. However, the evaluation is inadequate due to four major weaknesses described below. Despite these limitations, the authors conclude that current general-purpose LLMs lack adequate accuracy, which is already widely recognized. Its key contribution should instead be to provide concrete recommendations for the development of specialized LLMs for this task, which is completely absent. Developing such specific LLMs would be highly valuable, as they could substantially reduce the time required by researchers to analyze signaling networks.

Strengths

The manuscript raises a good question: whether current LLMs can accurately generate signaling networks from gene lists.

Weaknesses:

(1) The authors evaluate LLM performance using only three signaling networks: "hypertrophy", "fibroblast", and "mechanosignaling". Given the large number of well-established signaling pathways available, this is not a comprehensive assessment. Moreover, the analysis need not be restricted to signaling networks. Other network types, including metabolic and transcriptional regulatory networks, are already accessible in well-known databases such as KEGG, Reactome, BioCyc, WikiPathways, and Pathway Commons. Including these additional networks would substantially strengthen the evaluation.

We agree with the reviewer that our evaluation of LLM performance is not comprehensive of all signaling networks, and that the benchmarking was previously limited to signaling networks. The purpose of this study is to benchmark LLMs against peer-reviewed computational models that make testable predictions and are highly validated experimentally, of which these three signaling networks are strong examples. KEGG, Reactome, Biocyc, WikiPathways, and Pathway Commons are databases that contain collections of individual reactions or pathways but are not computational models themselves and have not been experimentally validated in that sense.

While this study focuses primarily on signaling networks, we agree that it would be useful to evaluate how well LLMs perform in generating networks of another type, for which predictive and validated computational models are available. Therefore, in new Figure 3 we now test the ability of LLMs to reconstruct the E. Coli core metabolic network, as well as test its ability to predict growth on metabolic substrates using flux balance analysis. We find that Claude Opus 4.6, GPT 5.2 Pro, and Gemini 3 Pro Preview perform well at reconstructing reactions from E. Coli core metabolism, but these reactions are not sufficient to predict growth on a variety of substrates. Given this expansion of scope, we replaced “signaling networks” in the title with “biochemical networks”.

(2) In LLM evaluation, the authors use the gene lists that exactly match those in their "ground truth" networks, thereby fixing the set of nodes and evaluating only the predicted edges. However, in practical research, the relevant genes or nodes are not fully known. A more realistic assessment would therefore include gene lists with both genes present in the ground-truth network and additional genes absent from it, to evaluate the ability of the LLM to exclude irrelevant genes.

We agree with the reviewer that evaluating the capacity of these LLMs to exclude additional genes is interesting. But because biological networks are always incompletely known, there is no “ground truth” of genes absent from a given network. Therefore, for the most rigorous benchmarking against a “ground truth”, we examine the positive predictive ability of LLMs. However, in response to this comment and point 3 below, we further examine additional measures of performance that include “false positives”.

(3) The authors report only the recall/sensitivity of the LLM, without assessing specificity. In practical applications, if an LLM generates a large number of incorrect interactions that greatly exceed the correct ones, researchers may be misled or may lose confidence in the LLM output. Therefore, a comprehensive evaluation must include both sensitivity and specificity. Furthermore, it would be informative to check whether some of the "false positives" might in fact represent biologically plausible interactions that are absent from the manually curated "ground truth". Manually generated "ground truth" can overlook genuine interactions, and the ability of LLMs to recover such missing edges could be particularly valuable. This may even represent one of the most important potential contributions of LLMs.

We agree with the reviewer that additional metrics could inform the evaluation of LLM performance. Therefore, as recommended, we calculated sensitivity, specificity, precision, negative predictive value (NPV), accuracy, and F1 score for each of the network models (hypertrophy, fibroblast, and mechanosignaling). These new results are summarized in confusion matrices shown in a new Supplementary Figures 3, 4, and 5. We performed this additional benchmarking across the 10 replicates for each LLM.

One limitation of this approach is the substantial class imbalance within these confusion matrices. Because we interpret actual negatives as connections that are not found between any nodes of the ground truth models, there will be >10x more true negatives than any of the other classes. This makes specificity, accuracy, and the negative predictive value less informative.

To illustrate this point, consider the ground-truth hypertrophy network which contains 191 connections between 106 nodes. The total number of possible connections between any two nodes is 1062 = 11,236. Given that there are 191 actual positives, that leaves 11,045, actual negatives (as illustrated in the null predictor of Supplementary Figure 3A). As the number of node-to-node connections predicted by LLMs is on the order of a few hundred, the number of true negatives is always in the thousands, often outweighing the true positives in the specificity or accuracy calculations.

The effects of this class imbalance are illustrated with the results of the “null predictors” in Supplementary Figure 4A, which are hypothetical models that fail to predict any connection between nodes (have predicted positive values of 0). These null predictors have sensitivities of 0 and specificities of 1 and high accuracies and NPVs because of the high true negative rates.

The precision and F1 scores calculated using these confusion matrices are robust to these class imbalances. Indeed, there is substantial heterogeneity in the number of “false positives” connections generated by the LLMs as illustrated by the precision and F1 scores. We are hesitant to unequivocally label these novel, predicted connections as false positives because, as the reviewer points out, these connections could represent true molecular interactions that were undiscovered at the time of the ground truth models’ conception but have since been demonstrated experimentally and published.

(4) It is widely known that applying differential equation models to highly complex biological networks, such as the three networks in the manuscript, is meaningless, because these systems involve a large number of parameters whose values can drastically alter the results. As Richard Feynman once said: "with four parameters I can fit an elephant, and with five I can make him wiggle his trunk." Thus, the evaluation of LLMs on "logic-based differential equation models" does not make much sense.

Differential equation models have been the primary mathematical framework for studying complex systems for decades. We refer readers unfamiliar with differential equations to the Nobel prize-winning work of Hodgkin/Huxley (Physiology or Medicine1963, action potential of neurons), Prigogine (Chemistry 1977, non-equilibrium thermodynamics and pattern formation), John Nash (Economics 1994, game theory and Nash Equilibrium), Merton/Scholes (Economics 1997, dynamics of financial derivatives), and Manabe/Hasselmann (Physics 2021, dynamic modeling of atmosphere and oceans).

We are confused by the quote of a joke by Richard Feynman about fitting equations to data in the shape of an elephant. While this famous joke is amusing, it is both misattributed (it was a recollection by Enrico Fermi in 1953 of a joke once made by John von Neumann) and deliberately hyperbolic (see https://en.wikipedia.org/wiki/Von_Neumann%27s_elephant). Regardless, the relevance of this joke to our study is unclear, because we are not fitting equations to data. As described in the text, in previous studies we validated the predictions of these three logic-based network models with experimental data that was not used to develop the models.

Reviewer #1 (Recommendations for the authors):

(1) All figures are in very poor resolution.

Thank you for identifying this. We have fixed this issue, which was due to SVG embedding. We now embed as higher resolution PNG and provide full resolution files separately.

(2) The manuscript does not include data availability or code availability.

As described in the Methods, all code and data is now available via GitHub (https://github.com/saucermanlab/LLM-network-generation).

Reviewer #2 (Public review):

(1) Information on the accuracy of directionality of interaction would help understand if there is a bias towards either a positive or negative association.

To assess if LLMs are biased in predicting either stimulatory or inhibitory connections, we examined the proportion of stimulatory and inhibitory connections for each ground truth model along with the prediction sets from the different LLMs (Author response table 1). These findings suggest that any directional bias is minimal and not conserved across the different ground truth models. These tables were not included in the revised manuscript.

The primary subset of reaction types that the LLM’s tend to miss are often downstream, cell-type specific, and/or gene regulatory connection (e.g. Figure 1B). This observation is consistent with the fact that the ground truth models were constructed using experimental evidence from specific publications involving defined experimental models and cell types whereas the LLM’s presumably draw from the entire corpus of published literature.

(2) Do all LLMs capture similar information, or are some LLMs better at capturing certain information than others? Further to this, it would be interesting to look into whether amalgamating information across all three LLMs results in a more accurate network.

The reviewer asks interesting questions that can be qualitatively answered in Figure 1B and in the network visualizations (Supplementary Figures 1 and 2). These diagrams illustrate predicted connections that are shared between the nodes. To include a more quantitative, comprehensive assessment of this overlap, we have included Venn diagrams showing the extent to which LLMs capture shared information (Supplementary Figures 3-5). One such Venn diagram for the hypertrophy model is included in Supplementary Figure 3C. Indeed, it seems that there is substantial overlap in the “false positive” connections predicted by Claude and GPT. It could be interesting to evaluate if these connections are reflective of newly discovered molecular interactions, as referenced in our response to reviewer one point 3.

While amalgamating information across all three LLMs might generate a more accurate network, these Venn diagrams illustrate that there remain connections within the ground truth models that are not represented in any of the prediction sets from the LLMs. Indeed, the union of all predicted connections for the hypertrophy network made by any of the 10 replicates from the different LLMs would still lack 33 ground truth connections (Supplemental Figure 3C).

(3) Would it be possible to retrieve a confidence value of the interactions from the LLMs and conduct Precision, recall rate, AUPR and calibration analyses? These metrics would also help with the benchmarking process.

We agree with the reviewer that including additional evaluative metrics is instructive. Therefore, for each reference network, we have included precision, specificity, recall, accuracy, and F1 score calculations (Supplementary Figures 3-5). Additionally, we conducted calibration analyses to illustrate the LLM’s reliability. While this latter analysis is interesting, we note that the calibration curves were generated using the frequency of predictions among the 10 replicates as a proxy for confidence. This frequency is highly sensitive to the temperature parameter of the LLMs. Different temperature settings could substantially influence the trajectories of these confidence curves and the distributions of the associated histograms. See Supplemental Figure 3B for calibration analyses of the hypertrophy model.

We considered performing analyses resembling AUPR as suggested by the reviewer, but we did see a defensible way to vary a “threshold” for the classification of a connection as positive or negative. In this study, a connection predicted by the LLM either does or does not exist within the ground truth mode.

Typos:

(1) Generated networks to predict THE CLASSIC "fetal gene program" gene expression.

Thank you for catching this error. We have corrected it.

(2) A manually curated network has A functional accuracy of

Thank you for catching this error. We have corrected it.

Version 2

Abstract

Computational models of biochemical networks provide frameworks for predicting how molecular cues guide cell decisions. These models are typically limited by the time-intensive manual curation required to extract network mechanisms from incomplete literature. Here, we test whether general-purpose large language models (LLMs) can generate accurate models of signaling and metabolic networks. We find that general-purpose LLMs generate 24-65% of the reactions of literature-curated signaling networks for cardiomyocyte hypertrophy, myofibroblast activation, and mechanosignaling. Further, logic-based models based on these networks predict responses to perturbations with accuracies of 6-33%. In the context of metabolic modeling, LLMs are able to generate 64-91% of the reactions within the core Escherichia coli metabolic network and demonstrate highly variable accuracies in predicting substrate utilization. Current general-purpose LLMs generate biochemical networks with moderate accuracy, and this study provides a pipeline and benchmarks to guide future improvements.

Main text

The ability of cells to grow or respond to external stimuli is fundamental to life, mediated by complex metabolic and signaling networks. Computational models of signaling networks have been widely used to understand their systems properties and guide new experiments1, but they typically require substantial manual literature curation of network reactions and rate constants2. Similarly, metabolic network models provide a powerful framework for predicting cellular metabolism and substrate utilization, but their construction has historically required manual curation of stoichiometric coefficients, reaction reversibility, and metabolite–reaction connectivity.

Artificial intelligence (AI), and large language models (LLMs) in particular, have demonstrated strong performance in tasks that traditionally required labor-intensive knowledge curation. For example, general-purpose LLMs have scored highly on the U.S. Medical Licensing Examination3, law exams4, and Advanced Placement exams4. AI has also made substantial contributions in biology, including predicting protein structure5 and cell types6, and is expected to play a central role in the development of comprehensive virtual cells7,8. However, to our knowledge, LLM-generated predictive models of biochemical networks have not been reported or benchmarked. Here, we develop pipelines for LLM-generated models of signaling and metabolic networks, and we test whether general-purpose LLMs accurately predict network structure. Additionally, we assess the LLM-generated networks’ accuracy in predicting perturbation responses, as benchmarked against manually curated models and experimental evidence9–11.

We first established a computational pipeline to generate models of signaling networks from LLMs using a specified gene set (Figure 1A). For benchmarking, each gene set was derived from annotations of three manually curated, logic-based network models that have been extensively validated using experimental data9–11. We optimized a programmatic prompt that iteratively queries an LLM for directed interactions (activation or inhibition) among the gene products in the context of a particular phenotype (here “hypertrophy”, “fibroblast”, or “mechanosignaling”). We then extracted the returned directed interactions, examined network structure, and converted the network into a logic-based formalism to simulate network responses to perturbations12. Each LLM was prompted 10 times per network to examine reproducibility (see methods for prompting parameters).

Signaling networks generated by general-purpose large language models.

A) Schematic of the pipeline for LLM-generated models of signaling networks. B) Network reactions recalled by three large language models (Gemini 3.0, orange; GPT5.2, blue; Claude4.6, green) compared with a “Ground Truth” literature-curated and validated cardiomyocyte hypertrophy signaling network9 (gray reactions). Edge color intensities indicate how frequently each reaction was predicted among the replicates (n = 10). LLM networks were generated using iterative prompts based on the gene set of the Ground Truth hypertrophy network. C–E) Summary of reaction recall for three literature-curated signaling networks (C, hypertrophy9; D, fibroblast10; and xsE, mechanosignaling11) by Gemini, GPT, and Claude. * Indicates p < 10−9 in one-sample t-test between LLM-generated replicates and the ground truth network.

Leveraging this pipeline, we tested the extent to which three general-purpose LLMs (GPT-5.2 Pro, OpenAI; Gemini 3.0 Pro-Preview, Google; and Claude Opus 4.6, Anthropic) could reconstruct the structure of the manually curated cardiomyocyte hypertrophy signaling network9, based on its gene annotations alone.

All three LLMs captured aspects of this manually curated network, but with varying coverage depending on the region of the network and the specific LLM (Figure 1B). In particular, GPT 5.2 Pro, Gemini 3.0 Pro-preview, and Claude Opus 4.6 all performed relatively well in reconstructing upstream ligand-receptor interactions (e.g. EGF to EGFR) and well-established second-messenger cascades such as the β-adrenergic receptor–cAMP–CREB or phospholipase C–diacylglycerol–PKC pathways. They also accurately predicted the structure of highly conserved signaling axes such as PI3K-Akt-mTor and calcium-calmodulin-calcineurin-NFAT. In contrast, the three LLMs were less able to reconstruct regulation of downstream transcription factors (e.g. cFOS, NFAT, cJUN, MEF2) and gene expression (SERCA, αMHC, βMHC, natriuretic peptides ANP/BNP) or cell area / hypertrophy. The lower reconstruction accuracy of downstream signaling may arise from cardiomyocyte-specific reactions. Overall, the three LLMs reconstructed 37.70-63.87% of network reactions in the cardiomyocyte hypertrophy network (Figure 1C).

To what extent is the limited recall of LLM-generated networks specific to cardiomyocyte hypertrophy? To address this question, we used the same reasoning models to reconstruct network models of fibroblast activation10 (91 nodes, 134 reactions) and mechanosignaling11 (94 nodes, 125 reactions). Like the hypertrophy model, these networks were previously manually reconstructed from the literature, translated into predictive logic-based models, and had their predictions validated against substantial experiments from the literature not used to develop the model. Consistent with the hypertrophy network, LLM-generated models of fibroblast activation and mechanosignaling reconstructed 24.07-60.49% and 29.07-64.53% of their manually curated counterparts (Figure 1D and E).

We then asked whether the regional coverage of LLM-generated networks of fibroblasts and mechanosignaling mirrors that of cardiomyocytes. Indeed, the LLMs consistently reconstructed ligand-receptor interactions and highly conserved pathways such as cAMP and Ras/Raf/ERK, but did not consistently recall AT1R signaling, PDGFR-abl signaling, and downstream regulation of transcription factors, genes, and extracellular matrix (e.g. MMP-collagen subnetwork) (Supplementary Figure 1). Unlike hypertrophy and fibroblast networks, mechanosignaling is mediated by stretch-dependent proteins that are relatively less characterized. Indeed, the LLM-generated networks for mechanosignaling did not adequately enumerate direct mechanosensory reactions and also had limited coverage of gene regulation (Supplementary Figure 2). Still, core conserved pathways for calcium regulation, regulation of protein synthesis, and MAPK signaling were reconstructed robustly.

Models of signaling networks are useful not just for their structure, but also for simulating the response to new perturbations. Prior to benchmarking, the hypertrophy, fibrosis, and mechanosignaling networks were automatically translated into logic-based differential equation models using the Netflux framework12,13. This allows network-wide prediction of dynamics in response to desired perturbations and does not require prior knowledge of rate constants or concentrations13. The predictions of these three logic-based models have been rigorously validated against experimental evidence not used to develop the model9–11,14. Using the same Netflux approach, we generated logic-based models based on all of the LLM-generated network structures shown in Figure 1.

Returning our focus to the hypertrophy network, we examined the ability of LLM-generated networks to predict the classic “fetal gene program” gene expression signature in cardiomyocytes. The manually curated model accurately predicts that AngII and/or ISO increase the expression of ANP, BNP, GATA4, βMHC, sACT, and increase cell area/size, while decreasing expression of SERCA and αMHC. In contrast, the LLM-generated network models produce limited aspects of this signature, with the best performance in predicting ISO-dependent ANP in the GPT version (Figure 2A). We then tested whether this validation accuracy scales across all available manually curated experiments that cover a wider array of perturbations and measured outputs. Overall, for 114 validations where the manually curated network has a functional accuracy of 94.74%, LLM-generated networks have an accuracy of 6.14-33.33% (Figure 2B). We then repeated this systematic experimental validation in comparison to literature-based models of the fibroblast (82% validation against independent literature) and mechanosignaling (78% validation). LLM-generated models of the fibroblast (Figure 2C) and mechanosignaling (Figure 2D) networks exhibited validation accuracies similar to the hypertrophy model: 16.87-21.69% and 5.81-14.53%, respectively. Across all three networks, validation accuracy of LLM-generated models for perturbation responses is lower than their accuracy for reconstructing network structure.

Experimental validation of perturbation responses predicted by LLM-generated signaling network models.

A) Representative validations of network models generated by manual curation or by LLMs (Gemini, GPT, Claude), in comparison to experiments in conditions of Angiotensin II (AngII) or isoproterenol (ISO) from the literature9. B–D) Summary of systematic validations of manually curated (Ground Truth) and LLM-generated reconstructions of hypertrophy, fibroblast, and mechanosignaling network models against perturbation experiments from the literature (n = 114, 83, and 171 experiments, respectively). * Indicates p < 10−11 in one-sample T test between LLM-generated model validation scores (n = 10 replicates) and ground truth model validation accuracy.

It is important to note that the functional validation simulations used LLM-generated signaling network models after pruning to the subset of reactions contained in the ground truth networks. This enabled direct mapping of LLM-predicted interactions onto the “AND” formalism used in the logic-based network models, but also excluded predicted biological interactions that were generated by the LLMs but absent from the curated networks, which we tentatively classify as false positives (LLM+, ground-truth−). When this constraint was removed and the full set of LLM-predicted interactions was used for functional validation, model performance did not substantially improve, except for the fibroblast network (Supplementary Figure 6A-C). Additionally, the average size of the unconstrained LLM-generated networks was substantially larger than the manually curated network (Supplementary Figure 6D).

Additional performance metrics, including confusion matrices and calibration analyses for each signaling network, are provided in Supplementary Figures 3–5. Notably, Gemini generally exhibited substantially higher precision than Claude or GPT in predicting signaling network reactions. F1 scores were relatively consistent among the evaluated LLMs.

Reactions generated by the LLMs but absent from the ground-truth signaling networks were classified as “false positives” for the purpose of calculating performance metrics. This classification should be interpreted cautiously, as some LLM-predicted/ground-truth-negative reactions may represent true biomolecular interactions or regulatory relationships that were not captured in the manually curated reference models. To assess the impact of retaining these additional predictions, we evaluated the functional performance of the full LLM-generated signaling networks without pruning LLM-predicted/ground-truth-negative reactions. This analysis revealed only marginal improvements, and in some cases reduced accuracy, for the hypertrophy and mechanosignaling networks (Supplementary Figure 6A and C). In contrast, the unrefined fibroblast network models exhibited substantially greater functional validation performance than the pruned reconstructions, which retained only LLM-predicted/ground-truth-positive reactions (Supplementary Figure 6B). The Claude- and GPT-generated signaling networks contained more connections than the corresponding manually curated networks – this increased network density may have contributed to their improved functional performance within the fibroblast context (Supplementary Figure 6D).

Expanding the scope of our LLM-benchmarking, we asked how effectively these models could reconstruct a metabolic network of core Escherichia coli (E. coli) metabolism15. The reconstruction was guided by the genes represented in the gene– protein–reaction (GPR) associations of the reference network. We adapted our prompting approach to generate the information needed to define metabolic reactions, including substrates, products, stoichiometry, and reversibility. These outputs were used to assemble LLM-generated metabolic network models for comparison against the manually curated reference network, as described in the Methods.

There is notable heterogeneity in the regional coverage of this metabolic network by the different LLMs. Generally, both Claude and Gemini were able to accurately reconstruct the glycolysis and tricarboxylic acid cycle (TCA) while GPT failed to capture a few specific reactions in each process including the phosphoglycerate mutase and kinase (PGM and PGK respectively). Additionally, GPT failed to reconstruct the lactate dehydrogenase and lactate transport reactions. Gemini and Claude performed poorly in recreating the malate and fumarate transport reactions feeding into the TCA cycle. Overall, the LLM-generated metabolic network models demonstrated markedly greater coverage of the ground-truth core E. coli metabolic network than the predicted signaling networks achieved for their respective ground-truth networks. Specifically, the LLM-generated models reconstructed 63.63–90.90% of reactions within the manually curated core E. coli network (Figure 3C).

Despite substantial improvement in coverage of the ground truth reactions, the LLM-generated metabolic networks models largely underperformed in the substrate utilization analysis. Indeed, the majority of the networks yielded infeasible flux solutions or predicted no growth on all of the substrates (Figure 3D). Interestingly, a sparse subset of the Gemini- and Claude-generated models achieved strong substrate utilization prediction performance, with accuracies approaching the score of the ground-truth model (Figure 3E).

Metabolic network models generated by general-purpose LLMs and substrate utilization analysis.

A) Schematic of pipeline for LLM-generated models of the core E. coli metabolic network. B) Network reactions recalled by three large language models (Gemini, orange; GPT, blue; Claude, green) compared with a “Ground Truth” manually curated metabolic network. Bar charts illustrate how frequently each reaction was predicted by the different LLMs in the replicates (n = 10). Inset shows zoomed-in model coverage of the ADK1 reaction. LLM networks were generated using iterative prompts based on the GPR gene list of the core E. coli metabolic network. C) Summary of reaction recall accuracy by Gemini, GPT, and Claude. D) Heatmap illustrating substrate utilization predictions of the ground truth model to and LLM-generated models compared to experimental data. Color intensity within the LLM columns indicates frequency of growth prediction amongst the 10 replicates. E) Summary of model performance compared to the experimental data across carbon sources. * Indicates p < 10−4 in one-sample T test between LLM-generated replicates (n = 10) and the ground truth network. Within the network visualization, diamonds indicate reactions while rectangles indicate metabolites.

In conclusion, we developed pipelines that use general-purpose LLMs to generate predictive models of signaling and metabolic networks. We benchmarked the signaling network pipeline across three LLMs against large-scale, manually curated, and experimentally validated network models. Across cardiac hypertrophy, fibroblast, and mechano-signaling networks, LLM-generated models exhibit 24.07-64.53% reconstruction of reactions from the manually curated signaling models. Reconstruction accuracy is high for ligand-receptor interactions and well-conserved pathways, but lower for cell-specific gene regulation or biophysical regulation.

Similarly, we benchmarked the LLMs’ performance in generating a metabolic network using experimental data and a manually curated reference model. Although the LLMs achieved greater structural recovery of the core E. coli metabolic network relative to the signaling networks, these reconstructions did not translate into robust functional performance as assessed by substrate utilization analysis. This discrepancy may reflect gaps in key reconstructed metabolic pathways and aligns with findings from the signaling network models, where incomplete reconstruction of critical pathways likely limited the accuracy of simulated perturbation responses. The ability of LLMs to reconstruct core E. coli metabolism may reflect the depth of existing literature on this model system, and similar approaches may perform worse for metabolic networks that are less extensively characterized.

This benchmarking does not yet test the ability of LLMs to reconstruct aspects of signaling that are not represented in literature-curated networks, such as isoform-specific regulation16 or liquid-liquid phase separation17. Such mechanisms are not yet adequately characterized for global inclusion in either literature-curated or LLM-generated models.

Future efforts could build on the pipelines developed in this work. Improving the performance of LLM-generated network models may require hybrid manual–LLM workflows or specialized AI models tailored to biological network reconstruction. Another promising direction would be to implement an agentic framework in which some agents propose plausible signaling reactions between genes, while others serve in a reviewer role to critique and approve candidate reactions. Such structured critique paradigms have shown promise in AI-oriented research contexts18. As LLMs and other AI models improve in their ability to synthesize and connect biological knowledge derived from experimental studies, developing rigorous platforms for organizing and evaluating this information will be increasingly important.

Methods

Prior knowledge models of signaling networks

We benchmarked LLM-generated network models using three large-scale network models based on manual curation of directed interactions using the literature9–11. These models contain a range of biological processes including ligand-receptor interactions, phosphorylation, gene regulation, and cellular phenotypes. Each node is annotated with corresponding genes. The models were previously implemented as systems of logic-based differential equations13, enabling validation of predictions to new perturbations with accuracies of 94.74% (hypertrophy), 81.93% (fibroblast), and 77.33% (mechanosignaling)9–11. Validation simulations were performed with parameter sets from each manually curated signaling network model’s original publication9–11.

Querying large language models to reconstruct signaling networks

We evaluated three widely used, general-purpose large language models (LLMs): GPT-5.2 Pro (OpenAI), Gemini 3.0 Pro-preview (Google), and Claude Opus 4.6 (Anthropic). For each model, prompting was performed through the corresponding application programming interface (API) using structured outputs to standardize response formatting across models. Temperature was set to 1.0, the maximum output length was set to 10,000 tokens, and 10 replicate outputs were generated for each LLM–signaling network pair, yielding 90 total outputs. The prompts consisted of three steps:

The list of gene symbols, along with the description: “List of genes and other signaling nodes:”.

Description of the desired outputs: “For the first {batch_size} entries in this list of genes, proteins, and other signaling nodes from a {desc} signaling network, please provide more than 0 but fewer than {max_connections} direct interactions with other nodes from the list supported by available literature. Return plausible interactions between these nodes in JSON format. Each interaction should have: source_node, target_node, and action (“stimulation” or “inhibition”).” Batch size was set at 20 genes, max connections set to the maximum number of connections exhibited in that parent network, and phenotype set at “cardiac hypertrophy”, “fibroblast”, or “mechanosignaling” corresponding to the three networks used for benchmarking.

Instructions to repeat the process until reaching the end of the gene list: “That looks great! Please do the same operation for the next {len(chunk)} nodes! Thank you”

Storing LLM-output in structured format and converting to logic-based models

Structured LLM outputs for the signaling networks were stored in JavaScript Object Notation (JSON) as interaction lists with three standardized fields: source_node, target_node, and action. The action field encoded the predicted regulatory relationship as either stimulation or inhibition. JSON interaction lists were converted into Netflux format for automated conversion into logic-based ordinary differential equation models12,13. These logic-based models were subsequently used for functional validation simulations as shown in Figure 2. In Figure 1, we evaluated restrictive reconstructions of the predicted signaling networks by retaining only LLM-predicted reactions that were also present in the corresponding ground-truth network. This approach excluded LLM-predicted/ground-truth-negative reactions, limiting the functional evaluation to true-positive reactions. This approach is in contrast with that of Supplementary Figure 6 which illustrates functional evaluation of unrefined, full LLM-generated networks.

Constructing confusion matrices using LLM-generated signaling network models

For each predicted signaling network model, we calculated sensitivity, specificity, precision, negative predictive value (NPV), accuracy, and F1 score relative to the corresponding ground-truth network. These metrics were summarized as confusion matrices for the hypertrophy, fibroblast, and mechanosignaling benchmarks (Supplementary Figures 3–5). Actual positives were defined as node-to-node connections present in the ground-truth network, whereas actual negatives were defined as all possible node-to-node connections among ground-truth nodes that were absent from the manually curated model.

Because the number of absent node-to-node connections greatly exceeded the number of curated reactions, these classification tasks were characterized by substantial class imbalance. For example, the hypertrophy network contained 191 curated connections among 106 nodes, yielding 11,236 possible directed node-to-node connections and 11,045 actual negatives. Consequently, true negatives dominated the confusion matrices, making specificity, accuracy, and NPV less informative than precision, sensitivity, and F1 score.

This effect was illustrated using null predictors, defined as hypothetical models that predicted no connections. These null predictors had sensitivity values of 0 and specificity values of 1, while still achieving high accuracy and NPV because of the large number of true negatives.

Calibration analyses and predicted model reaction unions

Calibration analyses were enabled by the generation of 10 replicates per LLM per signaling network model. The Scikit Learn calibration curve package was utilized to visualize the results of this analysis19. The Venn Diagrams contained in Supplementary Figures 3-5 illustrate the total reaction populations predicted by the LLMs (across all replicates) and their overlap with the ground truth model reaction set.

Querying large language models to reconstruct a metabolic network

The core E. coli model contains only central carbon metabolism15. This small-scale model (72 metabolites, 95 reactions, and 137 genes) contains reactions of glycolysis, the pentose phosphate shunt, the tricarboxylic acid cycle, the glyoxylate cycle, gluconeogenesis, anaplerotic reactions, the electron transport chain, fermentation, the transfer of reducing equivalents, and nitrogen metabolism. The model was downloaded from the BiGG Database (ID: e_coli_core), and the associated gene symbols were extracted20. Similar to the signaling network strategy, these genes were passed to the LLM using a sliding window approach with structured outputs (JSON format) and 10 replicates were generated per LLM. The LLM versions, temperature, maximum tokens, and other prompting parameters were identical to those used in generating signaling networks. The specific prompt is included below:

“For the first {batch_size} entries in this list of genes from an E. coli metabolic network, please return a structured list of metabolic reactions using the provided JSON schema:

Each interaction should have: a reaction_id, substrates (list of metabolite and coefficient pairs in BiGG format with compartment suffixes ‘_c’ or ‘_e’), products (list of metabolite and coefficient pairs in BiGG format with compartment suffixes ‘_c’ or ‘_e’), and reversible (boolean). If a gene catalyzes multiple reactions, please return each reaction separately.

[iteration completed]

That looks great! Please do the same operation for the next {len(chunk)} genes! Thank you”

Out of the 95 reactions in this core network, 26 are not associated with specific gene symbols and constitute transport, sink, demand, exchange, or spontaneous reactions. Additionally, three transport reactions (oxygen, carbon dioxide, and acetaldehyde) are associated with a placeholder gene ‘s0001’. Therefore, reaction-recovery analysis was performed using the 66 gene-associated reactions as the reference set.

Precise BiGG notation was requested for the predicted metabolite sets to facilitate mapping back to the ground truth model. Because the JSON schema stored substrate and product coefficients as positive magnitudes, substrate coefficients were multiplied by −1 when constructing stoichiometric matrices. Metabolite IDs were mapped to BiGG metabolite IDs using case-insensitive matching. This enabled generation of stoichiometric matrices for each replicate with matched metabolite IDs to the ground truth as rows and arbitrary reaction IDs (generated by the LLM but not formatted to BiGG notation) as columns. To identify the reactions predicted by the models, we leveraged the metabolite IDs and coefficients to perform vector comparison to the ground truth model’s columns (curated reactions). This vector-based matching was performed for all non-zero coefficient columns in the generated stoichiometric matrices. If the predicted model included non-zero coefficients for a set of metabolites that precisely matched a reaction column in the ground truth matrix (set of non-zero coefficients), that reaction was designated as ‘recovered.’ Metabolites that could not be mapped to BiGG IDs were excluded from strict matching and reversibility annotations were not considered for the recovery analysis.

Substrate utilization analysis using LLM-generated metabolic network models

To perform substrate utilization analysis, we leveraged an analogous approach to the signaling network validations. The predicted reactions that matched ground-truth reactions (i.e. ‘true-positive’ reactions) were combined into a reconstructed metabolic network model that enabled flux balance analysis (FBA). FBA, which underlies substrate utilization analysis, requires a sufficiently connected metabolic network to admit feasible flux solutions; thus, several of these predicted networks yielded infeasible solutions for particular substrates. Reconstructed networks were supplemented with exchange, demand, sink, biomass, ATP maintenance, and other non-gene-associated reactions to enable full FBA. The LLM-generated reversibility Boolean annotations were used to assign reaction bounds. Reactions with reversible = TRUE were assigned bounds of [-1000, 1000], whereas irreversible reactions were assigned bounds of [0, 1000] in agreement with BiGG conventions.

All of the carbon sources for the model were tested including glucose, fructose, acetate, succinate, malate, fumarate, ethanol, acetaldehyde, formate, lactate, pyruvate, alpha-ketoglutarate, glutamate, and glutamine. For each substrate condition, the corresponding exchange reaction lower bound was set to −10, while other tested carbon-source exchange reactions were closed. All other medium components were set according to the default medium in the BiGG model. The minimum threshold to define growth was 10−6. FBA was performed using COBRApy v0.30.0 and the biomass reaction was used as the objective21. Experimental data was taken from the EcoCyc database’s list of growth-media for E. coli substrain MG165522. This database contains consensus experimental findings (‘growth’, ‘weak growth’, ‘no growth’, or ‘inconclusive’) for E. coli MG1655’s growth at 37° C under aerobic conditions using isolated carbon sources. Two supplementary publications provided experimental evidence that unengineered E. coli cannot grow when ethanol is supplied as the sole carbon source23,24.

The code and LLM-generated predictions in JSON format are freely available on GitHub at: https://github.com/saucermanlab/LLM-network-generation